---
title: "Single-Cell RNA-Seq Quality Control"
author: "Bushra Khan"
output: html_document
---

# Introduction and Environment Setup

Before starting the analysis, I needed to ensure my environment was properly set up. Working with single-cell data in R requires a massive amount of dependencies, and I immediately ran into issues.

When I first tried to set up the environment, I struggled with installing `Seurat` and `SingleCellExperiment` because Bioconductor packages often have strict version requirements. I had to explicitly use `BiocManager::install("SingleCellExperiment")` to resolve a crashing issue downstream during annotation. 

Always ensure your `renv` is synced!

```{r setup, include=FALSE}
# Load required libraries
library(Seurat)
library(tidyverse)
```

# Step 1: Loading the Data

I downloaded the raw 10x Genomics PBMC dataset. The data is extremely noisy, so I immediately initialized a Seurat object. I set two baseline thresholds right out of the gate:
- `min.cells = 3`: Filtering out any genes that aren't expressed in at least 3 cells.
- `min.features = 200`: Filtering out any cells with fewer than 200 genes (these are almost certainly empty fluid droplets, not real cells).

```{r load_data}
mat_path <- "../data/pbmc3k/filtered_gene_bc_matrices/hg19"
counts   <- Read10X(data.dir = mat_path)

pbmc <- CreateSeuratObject(counts = counts, project = "pbmc3k",
                            min.cells = 3, min.features = 200)
```

# Step 2: Mitochondrial Quality Control

I needed to identify dying or stressed cells. When a cell ruptures, its cytoplasmic RNA leaks out, but the mitochondrial RNA remains trapped inside. Therefore, a high percentage of mitochondrial RNA indicates a dead cell. 

I used a regex pattern `^MT-` to calculate this percentage.

```{r mt_qc}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
```

# Step 3: Visualizing the Unfiltered Distributions

Before I cut the data, I wanted to visualize the distributions to see what I was dealing with. I generated Violin and Scatter plots, forcing custom aesthetics to make the metrics distinct.

```{r visualize}
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, 
        pt.size = 0.1, cols = "#3A86FF") +
  theme(axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(title = "Quality Control Distributions (Pre-Filter)", y = "Count / Percentage")
ggsave("../figures/01_qc_violin.png", width = 10, height = 5, bg = "white")

FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt", pt.size = 0.5, cols = "#FF006E") +
  labs(title = "Mitochondrial RNA vs Total RNA", 
       x = "Total RNA Molecules Detected (nCount)", 
       y = "Mitochondrial Percentage (%)") +
  theme(legend.position = "none")
ggsave("../figures/01_qc_scatter.png", width = 6, height = 5, bg = "white")
```

# Step 4: Applying the Biological Cutoffs

Based on the distributions above, I established my filtering thresholds:
1. `nFeature_RNA > 200`: Drops empty fluid droplets.
2. `nFeature_RNA < 2500`: Drops cell doublets/multiplets.
3. `percent.mt < 5`: Drops dead/dying cells.

```{r subset}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
```

# Step 5: Save Checkpoint

I saved this filtered object so I don't have to re-run QC every time I tweak the downstream clustering algorithms.

```{r save}
saveRDS(pbmc, "../results/01_pbmc_filtered.rds")
```
