---
title: "Single-Cell Normalization & Clustering"
author: "Bushra Khan"
output: html_document
---

# Introduction

With the dead cells and doublets removed from the dataset, the next step was to normalize the RNA counts and cluster the cells mathematically.

```{r setup, include=FALSE}
library(Seurat)
library(ggplot2)
```

# Step 1: Loading the Clean Data

I loaded the filtered PBMC dataset from the QC phase so I could begin the mathematical transformations.

```{r load}
pbmc <- readRDS("../results/01_pbmc_filtered.rds")
```

# Step 2: Normalization & Feature Selection

Raw RNA counts are highly skewed by sequencing depth (some cells randomly get sequenced much deeper than others). I normalized the data using `LogNormalize` (scaling by 10,000 to match standard transcripts-per-million rates).

```{r normalize}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 10000)
```

Out of ~20,000 genes in the human genome, most are just background noise. I ran a variance stabilizing transformation (VST) to isolate the top 2,000 most highly variable genes that actually drive biological differences between cell types.

```{r var_features}
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)
```

I then centered and scaled the data so that highly expressed genes wouldn't artificially dominate the downstream Principal Component Analysis (PCA).

```{r scale}
pbmc <- ScaleData(pbmc, features = rownames(pbmc))
```

# Step 3: Dimensionality Reduction (PCA)

A 2000-dimension matrix is computationally impossible to cluster effectively. I ran Principal Component Analysis (PCA) to compress the data down to linear components of maximum variance.

```{r pca}
pbmc <- RunPCA(pbmc, features = VariableFeatures(pbmc))
```

I generated an Elbow Plot to visualize the drop-off in variance. I visually inspected this plot and determined that the vast majority of biological signal was captured in the first 10 PCs.

```{r elbow}
ElbowPlot(pbmc, ndims = 30)
ggsave("../figures/02_elbow.png", width = 5, height = 4, bg = "white")
```

# Step 4: Manifold Approximation & Clustering

Using those top 10 PCs, I built a K-Nearest Neighbor (KNN) graph to map cell-to-cell proximities.

```{r neighbors}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
```

I applied the Louvain algorithm (resolution = 0.5) to group these neighbors into distinct biological clusters.

```{r clusters}
pbmc <- FindClusters(pbmc, resolution = 0.5)
```

Finally, I crushed the 10-dimensional space down to 2 dimensions using UMAP so I could physically see the clusters on a 2D plot.

```{r umap}
pbmc <- RunUMAP(pbmc, dims = 1:10)
```

# Step 5: Visualizing the Clusters

I generated the UMAP plot. To ensure the legend wasn't just raw numbers, I labeled them properly. I specifically injected `theme_classic()` to override Seurat's default boxed aesthetic, applying professional X/Y axis arrows and bold titles for presentation-quality output.

```{r visualize}
# Rename the cluster identities so they look professional in the legend
pbmc$seurat_clusters <- paste("Cluster", pbmc$seurat_clusters)
Idents(pbmc) <- "seurat_clusters"

DimPlot(pbmc, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.5) + 
  theme_classic() +
  labs(title = "Mathematical Clustering (UMAP)", 
       x = "UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  theme(
    axis.line = element_line(arrow = arrow(length = unit(0.3, "cm"), type = "closed")),
    plot.title = element_text(face = "bold", size = 14)
  )
ggsave("../figures/02_umap_clusters.png", width = 7, height = 6, bg = "white")
```

# Step 6: Save Checkpoint

I saved the fully clustered object to pass to the final annotation phase.

```{r save}
saveRDS(pbmc, "../results/02_pbmc_clustered.rds")
```
