---
title: "Single-Cell Biological Annotation"
author: "Bushra Khan"
output: html_document
---

# Introduction

In Phase 2, the cells were grouped mathematically into clusters, but they were still completely anonymous (Cluster 0, Cluster 1, etc.). In this final phase, my goal was to assign real biological identities (e.g., T Cells, B Cells, Monocytes) to these mathematical groupings.

```{r setup, include=FALSE}
library(Seurat)
library(SingleCellExperiment)
library(SingleR)
library(celldex)
library(tidyverse)
```

# Step 1: Loading the Clustered Data

I loaded the 2D-mapped, clustered PBMC object.

```{r load}
pbmc <- readRDS("../results/02_pbmc_clustered.rds")
```

# Step 2: Marker Gene Discovery (Differential Expression)

I ran a Wilcoxon Rank Sum test across every single cluster to find genes that are highly expressed in one cluster but turned off in all the others. These are the "biomarkers".

```{r markers}
markers <- FindAllMarkers(pbmc, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
write.csv(markers, "../results/03_marker_genes.csv", row.names = FALSE)
```

# Step 3: Visualizing the Biomarkers (Heatmap)

To prove my differential expression worked, I grabbed the top 5 most defining genes for each cluster. 

When I first plotted this, the columns (cells) for the smaller clusters were completely squished out of existence by the larger clusters. To fix this, I randomly downsampled the cells so that every cluster had a maximum of 50 cells in the plot, ensuring they all got equal visual weight.

```{r heatmap}
top5 <- markers %>% group_by(cluster) %>% slice_max(avg_log2FC, n = 5)

# Downsample to prevent small clusters from being crushed
pbmc_downsampled <- subset(pbmc, downsample = 50)

DoHeatmap(pbmc_downsampled, features = top5$gene, size = 4) +
  theme(axis.text.y = element_text(size = 8)) # Ensure gene names are readable
ggsave("../figures/03_marker_heatmap.png", width = 12, height = 12, bg = "white")
```

# Step 4: Automated Biological Annotation

Instead of manually guessing cell types by eyeballing marker genes (which is highly prone to human error and bias), I used the industry-standard `SingleR` package to algorithmically match my cells against the curated Human Primary Cell Atlas database.

```{r singler}
ref <- celldex::HumanPrimaryCellAtlasData()
sce <- as.SingleCellExperiment(pbmc)
pred <- SingleR(test = sce, ref = ref, labels = ref$label.main)
```

# Step 5: Final Cell Type Map

I mapped the algorithm's predictions back onto my original Seurat object.

When I first mapped the UMAP, the text labels were sitting directly on top of the dots, making it look incredibly messy and unprofessional. I injected `repel = TRUE` into `DimPlot()` to draw elegant lines pointing from the labels to the clusters.

```{r umap_final}
pbmc$cell_type <- pred$labels

DimPlot(pbmc, group.by = "cell_type", label = TRUE, repel = TRUE, label.size = 4, pt.size = 0.5) +
  theme_classic() +
  labs(title = "Biologically Annotated Cell Types", 
       x = "UMAP Dimension 1", 
       y = "UMAP Dimension 2") +
  theme(
    axis.line = element_line(arrow = arrow(length = unit(0.3, "cm"), type = "closed")),
    plot.title = element_text(face = "bold", size = 14)
  )
ggsave("../figures/03_umap_celltype.png", width = 8, height = 6, bg = "white")
```

# Step 6: Final Data Save

I finally saved the fully processed object. I generated both the full version and a highly compressed "slim" version so that I wouldn't hit out-of-memory errors on Posit Connect Cloud when deploying my Shiny Dashboard!

```{r save}
saveRDS(pbmc, "../results/03_pbmc_annotated.rds")

# Generate slim version for deployment
slim <- DietSeurat(pbmc, counts=TRUE, data=TRUE, scale.data=FALSE, dimreducs=c("umap"))
saveRDS(slim, "../results/03_pbmc_slim.rds")
```
